\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
% The amssymb package is required for the \square command
\usepackage{amssymb}
\usepackage{parskip}
% \usepackage{fontspec}
% \setmainfont{Calibri}


\begin{document}

% Page 1: Key Project Requirements Checklist
\section*{Master Project Requirements Checklist}
\begin{itemize}[label=$\square$]
  \item \textbf{Unique Topic:} A non-common, real-world topic must be selected.
  \item \textbf{Proposal Deadline:} The proposal must be submitted by April 11th.
  \item \textbf{Project Deadline:} The final project must be completed by May 6th.
  \item \textbf{Technology Stack:} A full-stack application must be developed using Python, Tkinter, and a database.
  \item \textbf{User Validation:} Validation is required at both the GUI level and the database control layer.
  \item \textbf{Error Checking:} Robust error handling must be implemented throughout the code.
  \item \textbf{Professional GUI:} A polished design with professional colors and layout is mandatory.
  \item \textbf{Database Depth:} The implementation of multiple tables and relational queries (pivots) is expected.
  \item \textbf{Academic Integrity:} No AI-generated code is permitted; all work must be original and scannable.
\end{itemize}

\newpage

% Page 2: The Proposal
\textbf{Introduction to Digital Technology} \\
Jihoo Choi \\
Period: 2 \\
\textbf{Master Project Proposal} \\
Date Submitted: \today{}

\vspace{1em}

\noindent \textbf{What I am managing or tracking} \\
A comprehensive management system for a Minecraft server community and its associated virtual economy is to be developed. Player data, including unique identifiers, whitelist status, and total playtime, are recorded within a relational database. Furthermore, in-game currency transactions are tracked to ensure financial stability within the virtual environment. 

The system is designed to manage complex relationships between multiple tables. One table is utilized for core player profiles, while a secondary table is dedicated to transaction logs, linking each record to a specific player ID. Relational queries are performed to generate reports on wealth distribution and player activity trends, ensuring that the database depth exceeds basic single-table implementations.

\vspace{1em}

\noindent \textbf{Why this is meaningful to me} \\
A significant interest in the technical administration and maintenance of Minecraft servers has been maintained over several projects. The challenges associated with community management and the prevention of economic inflation within a sandbox environment are viewed as excellent opportunities for software-based solutions. This project allows for the practical application of database management skills to a hobby that has been consistently pursued.

The development of this tool is considered meaningful because it addresses real-world problems found in digital community moderation. By creating a professional GUI to interface with complex relational data, a bridge is built between backend server administration and frontend usability. The mastery of these full-stack components is regarded as a vital milestone in technical proficiency.

\vfill

\noindent \textbf{Attestation:} By signing below, I attest that I will write my own code.  I will only allow others to help me understand my errors or generically perform a single specific task.  All my code will be uniquely my own.  My database queries and GUI construction will all be my own.  I am aware that all code submissions will be digitally scanned for similarities and any source codes I submit will be in a scannable colorized PDF to Microsoft Teams. I fully understand that this project is a major assessment, and that plagiarism will result in disciplinary action.

\vspace{1em}

\noindent I further attest that zero percentage of my code has been generated in the creation of my code under penalty of administrative processing for plagiarism.
 
\vspace{4em}

\noindent \rule{5cm}{0.4pt} \hfill \rule{3cm}{0.4pt} \\
Signed \hfill Date

\vspace{1em}

This must be printed and signed, then scanned for submission with original signature.

\end{document}